A Tale of Two Transects

### 2020-01-29 writing/transect_paper.Rmd
move more data munging to drake plan, work on heat trees

---
title: "A Tale of Two Transects"
subtitle: "Also Two Sequencing Technologies, and Two Amplicons"
author: 
  - Brendan Furneaux^1^,
    Mohammad Bahram^1^,
    Anna Rosling^3^,
    Nourou S. Yorou^2^,
    Martin Ryberg^1^
  - ^1^Program in Systematic Biology,
    Department of Organismal Biology,
    Uppsala University,
    Uppsala, Sweden
  - ^2^Research Unit in Tropical Mycology and Plant-Fungi Interactions,
    LEB,
    University of Parakou,
    Parakou, Benin
  - ^3^Program in Evolutionary Biology,
    Department of Ecology and Genetics,
    Uppsala University,
    Uppsala, Sweden
date: "`r format(Sys.time(), '%B %d, %Y')`"
bibliography:
  - "all.bib"
  - "R.bib"
csl: molecular-ecology.csl
output:
  bookdown::pdf_document2:
    fig_crop: yes
    keep_tex: yes
    latex_engine: lualatex
    includes:
      in_header: preamble.tex
  bookdown::word_document2: 
    keep_md: yes
---

```{r setup, include=FALSE}
library("knitr")
library("tikzDevice")
library("ggplot2")
library("magrittr")
library("tidyverse")
library("drake")
library("here")
library("phyloseq")
tryCatch(
  tinytex::tinytex_root(),
  error = function(e) tinytex::install_tinytex()
  )
knitr::opts_chunk$set(
  dev = "pdf",
  # dev.args = list(pointsize=32),
  echo = FALSE,
  # results = 'asis',
  # fig.show = 'asis',
  fig.width = 6,
  fig.height = 4.5,
  message = FALSE,
  warning = FALSE,
  external = TRUE
)
options(
  tikzDefaultEngine = "luatex",
  tikzXelatexPackages = c(options("tikzXelatexPackages")$tikzXelatexPackages,
                          "\\usepackage[version=4]{mhchem}\n",
                          "\\usepackage{siunitx}\n",
                          "\\DeclareSIUnit\\year{a}\n",
                          "\\usepackage{fontspec}"),
  tikzMetricsDictionary = here("writing/transect_paper-tikzDictionary")
)

@@

reads_table <- cache$list() %>%
  purrr::keep(startsWith, "big_seq_table_") %>%
  purrr::map(readd, character_only = TRUE, cache = cache) %>%
  purrr::map(tibble::as_tibble, rownames = "sample") %>%
  purrr::map_dfr(tidyr::pivot_longer, -"sample", names_to = "seq", values_to = "reads") %>%
  dplyr::filter(reads > 0) %>%
  tidyr::extract(
    "sample",
    c("seq_run", "plate", "well", "region"),
    regex = "([pi][bs]_\\d{3})_(\\d{3})_([A-H]1?\\d)_(.+)"
  ) %>%
  # dplyr::mutate_at(
  #   "region",
  #   factor,
  #   levels = c("ITS1", "5_8S", "ITS2", "LSU1", "D1", "LSU2", "D2", "LSU3",
  #              "D3", "LSU4", "ITS", "LSU", "32S", "long")) %>%
  dplyr::group_by(seq_run, region, seq) %>%
  dplyr::summarize(reads = sum(reads)) %>%
  dplyr::mutate(length = nchar(seq)) %>%
  dplyr::ungroup()

region_table <- reads_table %>%
  dplyr::group_by(seq_run, region) %>%
  dplyr::summarize(
    ASVs = dplyr::n_distinct(seq),
    length_min = min(length),
    length_q1 = reldist::wtd.quantile(length, 0.25, weight = reads),
    length_med = reldist::wtd.quantile(length, 0.5, weight = reads),
    length_mean = weighted.mean(length, w = reads),
    length_q3 = reldist::wtd.quantile(length, 0.75, weight = reads),
    length_max = max(length),
    reads = sum(reads)
  )
loadd(big_seq_table_ITS2, cache = cache)
asv_table <- big_seq_table_ITS2 %>%
  tibble::as_tibble(rownames = "filename") %>%
  tidyr::gather(key = "seq", value = "reads", -1) %>%
  dplyr::filter(reads >= 1) %>%
  tidyr::extract(col = "filename",
                 into = c("seq.run", "plate", "well", "dir", "region"),
                 regex = "([a-z]+_\\d+)_(\\d+)_([A-H]1?[0-9])([fr]?)_([:alnum:]+).+") %>%
  dplyr::group_by(seq.run, seq) %>%
  dplyr::summarize(reads = sum(reads)) %>%
  tidyr::spread(key = seq.run, value = reads, fill = 0)

otu <- read_tsv(
  here("data/clusters/ITS2.table"),
  col_types = cols(
    .default = col_integer(),
    `#OTU ID` = col_character()
  )
) %>%
  column_to_rownames("#OTU ID") %>%
  as.matrix() %>%
  t() %>%
  as_tibble(rownames = "sample") %>%
  tidyr::extract(
    col = "sample",
    into = c("seq_run", "plate", "well"),
    regex = "([a-z]{2}_\\d{3})_(\\d{3})([A-H]1?\\d)"
  ) %>%
  tidyr::gather(key = "seq", value = "reads", -(1:3)) %>%
  dplyr::filter(reads >= 1) %>%
  dplyr::group_by(seq_run, seq) %>%
  dplyr::summarize(reads = sum(reads)) %>%
  tidyr::spread(key = seq_run, value = reads, fill = 0)

@@

```{r length-qual}

parse_qstat <- function(d) {
  mutate_at(d, "file", basename) %>%
  tidyr::extract(col = "file", into = c("seq_run", "plate", "well"),
                 regex = "([pi][bs]_\\d{3})_(\\d{3})-?([A-H]1?\\d)?[rf]?[.].*")
}

demuxlength <- readd(qstats_length, cache = cache) %>%
  parse_qstat() %>%
  filter(is.na(step), !is.na(well)) %>%
  group_by(seq_run) %>%
  summarize(length = reldist::wtd.quantile(length, weight = nreads)) %>%
  deframe()

demuxqual <- readd(qstats_erate, cache = cache) %>%
  parse_qstat() %>%
  filter(is.na(step), !is.na(well)) %>%
  group_by(seq_run) %>%
  summarize(qual = round(weighted.mean(erate, w = nreads, na.rm = TRUE), 4)) %>%
  deframe()

readcounts <- 
  readd(qstats_n, cache = cache) %>%
  parse_qstat()

rawcounts <- readcounts %>%
  filter(is.na(step), is.na(well)) %>%
  group_by(seq_run) %>%
  summarize(nreads =  prettyNum(sum(nreads), big.mark = " ")) %>%
  deframe()

demuxcounts <- readcounts %>%
  filter(is.na(step), !is.na(well)) %>%
  group_by(seq_run) %>%
  summarize(nreads = prettyNum(sum(nreads), big.mark = " ")) %>%
  deframe()

regioncounts <- readcounts %>%
  filter(step == "lsux") %>%
  group_by(seq_run) %>%
  summarize(nreads = prettyNum(
    sum(nreads) %/% if (seq_run[1] == "pb_500") 14 else 4,
    big.mark = " ")
  ) %>%
  deframe()
```

@@

Taxonomic assignment was performed to genus level separately on the ITS and LSU regions using Unite/Warcup and RDP, respectively, as taxonomic references.
For each region/reference combination, taxonomy was assigned using three algorithms:
the RDP Naïve Bayesian Classifier (NBC) as implemented in DADA2;
SINTAX [@edgar2016a] as implemented in VSEARCH v2.9.1 [@rognes2016];
and IDTAXA [@murali2018].
Each ASV was thus given up to nine preliminary taxonomic assignments (three region/reference combinations $\times$ three algorithms).

@@

```{r bioinfo}
bioinf_table <- bind_rows(
  enframe(rawcounts, name = "seq_run", value = "reads") %>%
    mutate(
      reads = as.integer(gsub(" ", "", reads)),
      step = "Raw"
    ),
  enframe(demuxcounts, name = "seq_run", value = "reads") %>%
    mutate(
      reads = as.integer(gsub(" ", "", reads)),
      step = "Trim"
    ),
  enframe(regioncounts, name = "seq_run", value = "reads") %>%
    mutate(
      reads = as.integer(gsub(" ", "", reads)),
      step = "LSUx"
    ),
  filter(region_table, region == "ITS2") %>%
    select(seq_run, reads, step = region, ASVs),
  pivot_longer(asv_table, -1, names_to = "seq_run", values_to = "reads") %>%
    filter(reads > 0) %>%
    mutate_at("seq", chartr, old = "T", new = "U") %>%
    inner_join(
      select(readd(allseqs, cache = cache), seq = ITS2, long, short, ITS, LSU) %>%
        pivot_longer(-1, names_to = "region", values_to = "consensus") %>%
        filter(!is.na(consensus)) %>%
        select(-consensus) %>%
        unique(),
      by = "seq"
    ) %>%
    group_by(seq_run, region) %>%
    summarize(reads = sum(reads, na.rm = TRUE), ASVs = n()) %>%
    select(seq_run, reads, step = region, ASVs),
  # pivot_longer(otu, -1, names_to = "seq_run", values_to = "reads") %>%
  #   filter(reads > 0) %>%
  #   mutate_at("seq", chartr, old = "T", new = "U") %>%
  #   inner_join(
  #     select(readd(allseqs, cache = cache), seq = ITS2, long, short, ITS, LSU) %>%
  #       pivot_longer(-1, names_to = "region", values_to = "consensus") %>%
  #       filter(!is.na(consensus)) %>%
  #       select(-consensus) %>%
  #       unique(),
  #     by = "seq"
  #   ) %>%
  #   group_by(seq_run, region) %>%
  #   summarize(reads = sum(reads, na.rm = TRUE), OTUs = n()) %>%
  #   select(seq_run, reads, step = region, OTUs)
) %>%
  left_join(select(datasets, seq_run, tech, amplicon), by = "seq_run") %>%
  mutate(
    step = ifelse(tech == "Ion" & step == "Trim", "Raw", step),
  ) %>%
  select(tech, amplicon, step, reads, ASVs) %>%
  pivot_longer(c("reads", "ASVs"), names_to = "type", values_to = "count") %>%
  filter(!is.na(count)) %>%
  mutate(
    step = factor(step, levels = c("Raw", "Trim", "LSUx", "ITS2",
                                   "short", "ITS", "LSU", "long")),
    tech = factor(tech, levels = c("PacBio", "Ion Torrent", "Illumina")),
    amplicon = factor(amplicon, levels = c("Long", "Short")),
    type = factor(type, c("ASVs", "reads"))
  ) %>%
  arrange(tech, amplicon, type) %>%
  pivot_wider(
    id_cols = "step",
    names_from = c("tech", "amplicon", "type"),
    values_from = c("count")
    ) %>%
  arrange(step) %>%
  column_to_rownames("step")

options(knitr.kable.NA = '–')
bioinf_header <- tibble(name = names(bioinf_table)) %>%
  separate(name, into = c("tech", "amplicon", "type"), sep = "_") %>%
  mutate(tech = paste(tech, plyr::mapvalues(tech, datasets$tech, datasets$machine, FALSE)))
bioinf_table %>%
  set_names(gsub(".*_", "", names(.))) %>%
  kable(
    booktabs = TRUE,
    caption = "(ref:readcounts)",
    align = "rrrrrr",
    format.args = list(big.mark = " "),
    linesep = c(rep("", 3), rep("\\addlinespace", 2), rep("", 4)),
    escape = FALSE
    ) %>%
  kableExtra::add_header_above(
    header = c(
      " " = 1,
      bioinf_header$amplicon[c(TRUE, FALSE)] %>% set_names(rep(2, length(.)), .)
    )
  ) %>%
  kableExtra::add_header_above(
    header = c(
      " " = 1,
      rle(bioinf_header$tech) %$% set_names(lengths, values)
    )
  )
```

@@

```{r venn-table-asv}
venn_ASV <-
  asv_table %>%
  select(seq, pb_500, pb_483, is_057) %>%
  mutate_if(is.numeric, list(found = ~. > 0)) %>%
  group_by_at(vars(ends_with("_found"))) %>%
  summarize_if(is.numeric, list(ASVs = ~sum(. > 0), reads = sum)) %>%
  ungroup() %>%
  mutate_at(vars(ends_with("_found")), as.integer) %>%
  tidyr::unite(col = "set", 1:3, sep = "") %>%
  pivot_longer(
    -1,
    names_to = c("seq_run", "what"),
    values_to = "value",
    names_pattern = "([[:alpha:]]+[-._]\\d+)_(.+)"
  ) %>%
  pivot_wider(names_from = "what", values_from = "value") %>%
  group_by(seq_run) %>%
  mutate_if(is.numeric, list(frac = ~./sum(.))) %>%
  mutate_at(
    vars(ends_with("_frac")),
    formatC,
    digits = 2,
    format = "f",
    drop0trailing = FALSE
  ) %>%
  mutate_at(
    c("reads", "ASVs"),
    format
  ) %>%
  mutate(
    ASVs_frac = if_else(grepl(" +0", ASVs), "", ASVs_frac),
    ASVs = if_else(grepl(" +0", ASVs), "", ASVs),
    reads_frac = if_else(grepl(" +0", reads), "", reads_frac),
    reads = if_else(grepl(" +0", reads), "", reads)
  ) %>%
  select(set, seq_run, ASVs, ASVs_frac, reads, reads_frac) %>% {
    left_join(
      group_by(., set) %>%
        summarize(ASVs = as.integer(first(ASVs[ASVs != ""]))),

      select(., set, seq_run, ASV_frac = ASVs_frac, reads, frac = reads_frac) %>%
      pivot_wider(
        names_from = "seq_run",
        values_from = c("ASV_frac", "reads", "frac")
      )
    )
  } %>%
  mutate(n = str_count(set, "1")) %>%
  arrange(n, desc(set)) %>%
  select(set, ASVs, ends_with("pb_500"), ends_with("pb_483"), ends_with("is_057")) %>%
  column_to_rownames("set")

@@

```{r venn-table-otu}
venn_OTU <-
  otu %>%
  select(seq, pb_500, pb_483, is_057) %>%
  mutate_if(is.numeric, list(found = ~. > 0)) %>%
  group_by_at(vars(ends_with("_found"))) %>%
  summarize_if(is.numeric, list(OTUs = ~sum(. > 0), reads = sum)) %>%
  ungroup() %>%
  mutate_at(vars(ends_with("_found")), as.integer) %>%
  tidyr::unite(col = "set", 1:3, sep = "") %>%
  pivot_longer(
    -1,
    names_to = c("seq_run", "what"),
    values_to = "value",
    names_pattern = "([[:alpha:]]+[-._]\\d+)_(.+)"
  ) %>%
  pivot_wider(names_from = "what", values_from = "value") %>%
  group_by(seq_run) %>%
  mutate_if(is.numeric, list(frac = ~./sum(.))) %>%
  mutate_at(
    vars(ends_with("_frac")),
    formatC,
    digits = 2,
    format = "f",
    drop0trailing = FALSE
  ) %>%
  mutate_at(
    c("reads", "OTUs"),
    format
  ) %>%
  mutate(
    OTUs_frac = if_else(grepl(" +0", OTUs), "", OTUs_frac),
    OTUs = if_else(grepl(" +0", OTUs), "", OTUs),
    reads_frac = if_else(grepl(" +0", reads), "", reads_frac),
    reads = if_else(grepl(" +0", reads), "", reads)
  ) %>%
  select(set, seq_run, OTUs, OTUs_frac, reads, reads_frac) %>%  {
    left_join(
      group_by(., set) %>%
        summarize(OTUs = as.integer(first(OTUs[OTUs != ""]))),

      select(., set, seq_run, OTU_frac = OTUs_frac, reads, frac = reads_frac) %>%
      pivot_wider(
        names_from = "seq_run",
        values_from = c("OTU_frac", "reads", "frac")
      )
    )
  } %>%
  mutate(n = str_count(set, "1")) %>%
  arrange(n, desc(set)) %>%
  select(set, OTUs, ends_with("pb_500"), ends_with("pb_483"), ends_with("is_057")) %>%
  column_to_rownames("set")

@@

```{r pb-vs-it, fig.cap="(ref:readcap)", fig.height=8}
big_table <- bind_rows(
  mutate(asv_table, type = "ASV"),
  mutate(otu, type = "OTU")
) %>%
  select(-seq) %>%
  select(type, everything())

names(big_table)[-1] <-
  tibble(seq_run = names(big_table)[-1]) %>%
  left_join(datasets, by = "seq_run") %>%
  glue::glue_data("{tech} {machine} - {amplicon}")

multi_table <- crossing(
  x_var = factor(glue::glue_data(datasets, "{tech} {machine} - {amplicon}"), ordered = TRUE),
  y_var = factor(glue::glue_data(datasets, "{tech} {machine} - {amplicon}"), ordered = TRUE),
  type = c("ASV", "OTU")
) %>%
  filter(
    x_var < y_var,
    x_var %in% names(big_table),
    y_var %in% names(big_table)
  ) %>%
  mutate_all(as.character) %>%
  group_by(x_var, y_var, type) %>%
  group_modify(
    function(d, g) {
      big_table %>%
        filter(type == g$type) %>%
        select(x = !!g$x_var, y = !!g$y_var) %>%
        filter(x > 0 | y > 0)
    }
  )

comparisons <- multi_table %>%
  group_modify(
    function(d, g) {
      filter_all(d, ~. > 0) %>%
        mutate_all(log10) %>%
      {bind_cols(
      lm(y ~ x, .) %>%
        broom::glance(.),
      lm(y ~ offset(x), .) %>%
        broom::tidy(.) %>%
        select(term, estimate) %>%
        pivot_wider(names_from = term, values_from = estimate)
      )}
    }
  ) %>%
  mutate(
    label = paste("R^2 ==", formatC(r.squared, format = "f", digits = 2))
  )

# comparisons <- crossing(
#   x_var = factor(datasets$seq_run, ordered = TRUE),
#   y_var = factor(datasets$seq_run, ordered = TRUE),
#   type = c("ASV", "OTU")
# ) %>%
#   filter(y_var < x_var) %>%
#   mutate_all(as.character) %>%
#   group_by(x_var, y_var, type) %>%
#   group_modify(function(d, g) {
#     big_table %>%
#       filter(type == g$type) %>%
#       select(x = !!g$x_var, y = !!g$y_var) %>%
#       filter(.data[[g$x_var]] > 0 & .data[[g$y_var]] > 0) %>%
#       mutate_all(log10) %>%
#       lm() %>%
#       {bind_rows(
#       broom::glance(.),
#       broom::tidy(.) %>%
#         select(term, estimate) %>%
#         pivot_wider(names_from = term, values_from = estimate)
#       )} %>%
#       bind_cols(d, .)
#     }) %>%
#   mutate(
#     label = paste("R^2 ==", formatC(r.squared, format = "f", digits = 2))
#   )

@@

# p <- multi_table %>%
#   filter(x > 0, y > 0, type == "ASV") %>%
#   ggplot(aes(y = y, x = x)) +
#   geom_point(alpha = 0.2, shape = 1) +
#   scale_x_log10() +
#   scale_y_log10(limits = c(1, 1e4)) +
#   coord_equal() +
#   xlab(NULL) +
#   ylab(NULL) +
#   geom_rug(aes(y = y), sides = "l", alpha = 0.2,
#            data = multi_table %>% filter(x == 0, y > 0)) +
#   geom_rug(aes(x = x), sides = "b", alpha = 0.2,
#            data = multi_table %>% filter(y == 0, x > 0)) +
#   facet_grid(y_var ~ x_var, switch = "both", space = "free") +
#   stat_smooth(method = "loess", formula = y ~ x) +
#   geom_abline(aes(intercept = `(Intercept)`, slope = 1),
#               linetype = 2L, alpha = 0.5, data = comparisons) +
#   geom_text(
#     aes(label = label),
#     data = comparisons,
#     x = 1,
#     y = 3.5,
#     color = "black",
#     hjust = 0,
#     parse = TRUE,
#     inherit.aes = FALSE
#   ) +
#     theme(strip.placement = "outside", strip.background = element_blank())
# g <- ggplotGrob(p)
# g$grobs[g$layout$name %in% c("panel-1-2")] <- NULL
# g$layout <- g$layout[!(g$layout$name %in% c("panel-1-2")),]
# grid.newpage()
# grid.draw(g)

expand_seq_run <- function(labels) {
  lapply(
    labels,
    )
}

@@

read_comparison <- function(big_table, comparisons, x, y, xlab, ylab) {
  x_var <- enquo(x)
  y_var <- enquo(y)
  big_table %>%
  filter(!!x_var > 0, !!y_var > 0) %>%
  ggplot(aes(x = !!x_var, y = !!y_var)) +
  geom_point(alpha = 0.2, shape = 1) +
  scale_x_log10() +
  scale_y_log10(limits = c(1, 1e4)) +
  coord_equal() +
  xlab(xlab) +
  ylab(ylab) +
  geom_rug(aes(y = !!y_var, x = 1), sides = "l", alpha = 0.2,
           data = big_table %>% filter(!!x_var == 0, !!y_var > 0)) +
  geom_rug(aes(y = !!x_var), sides = "b", alpha = 0.2,
           data = big_table %>% filter(!!y_var == 0, !!x_var > 0)) +
  stat_smooth(method = "loess", formula = y ~ x) +
  geom_abline(aes(intercept = mean(log10((!!y_var)[(!!y_var) >= 1])) -
                    mean(log10((!!x_var)[(!!x_var) >= 1])), slope = 1),
              linetype = 2L, alpha = 0.5) +
  geom_text(
    aes(label = label),
    data = filter(comparisons, .data$x == quo_name(x_var), .data$y == quo_name(y_var)),
    x = 1,
    y = 3.5,
    color = "black",
    hjust = 0,
    parse = TRUE,
    inherit.aes = FALSE
  ) +
  facet_wrap(~type, strip.position = "right", scales = "free")
}
# ggpubr::ggarrange(
#   ncol = 1,
#   labels = "auto",
#   read_comparison(
#     big_table,comparisons,
#     is_057, pb_500,
#     "Ion Torrent Short Reads", "PacBio Long Reads"),
#   read_comparison(
#     big_table, comparisons,
#     is_057, pb_483,
#     "Ion Torrent Short Reads", "PacBio Short Reads"),
#   read_comparison(
#     big_table, comparisons,
#     pb_483, pb_500,
#     "PacBio Short Reads", "PacBio Long Reads")
# )
```

@@

@@

```{r}
taxon_reads <- 
  readd(proto_physeq, cache = cache) %>%
  phyloseq::merge_samples(group = "seq_run") %>%
  phyloseq::otu_table() %>%
  t() %>%
  {.@.Data} %>%
  as_tibble(rownames = "label") %>%
  mutate_if(is.numeric, ~./sum(.)) %>%
  pivot_longer(-1, names_to = "seq_run", values_to = "reads") %>%
  filter(reads > 0) %>%
  left_join(select(datasets, seq_run, amplicon), by = "seq_run") %>%
  expand_grid(
    tibble(
      reference = c("unite", "warcup", "rdp_train"),
      region = c("ITS", "ITS", "LSU")
    ),
      method = c("dada2", "idtaxa", "sintax")
  ) %>%
  filter(
    ifelse(amplicon == "Short", region == "ITS", TRUE),
    ifelse(amplicon == "Long", region %in% c("ITS", "LSU"), TRUE)
  ) %>%
  left_join(
    readd(taxon_table, cache = cache) %>%
      mutate(region = ifelse(region == "short", "ITS", region)) %>%
      mutate_at("taxon", na_if, "NA") %>%
      select(label, method, reference, rank, taxon, region) %>%
      pivot_wider(
        names_from = "rank",
        values_from = "taxon"
      ),
    by = c("label", "reference", "region", "method")
  ) %>%
  mutate(
    kingdom = ifelse(!is.na(phylum) & reference == "warcup", "Fungi", kingdom)
  )
```

(ref:taxa-kingdom) Taxonomic assignment of ASVs to kingdom.
N/A denotes ASVs which could not be identified to kingdom level.
Height of each bar represents the fraction of reads represented by that bar.

```{r taxa-kingdom, fig.cap = "(ref:taxa-kingdom)"}
taxon_reads %>%
  group_by(reference, method, seq_run, kingdom) %>%
  summarize(reads = sum(reads), ASVs = n()) %>%
  ggplot(aes(x = method, y = reads, fill = kingdom)) +
  geom_bar(position = "stack", stat = "identity", color = "white", size = 0.2) +
  facet_wrap(~seq_run + reference, nrow = 1, scales = "free_x") +
  theme(axis.text.x = element_text(angle = 90))
```

(ref:ASV-kingdom) Taxonomic assignment of ASVs to kingdom.
N/A denotes ASVs which could not be identified to kingdom level.
Height of each bar represents the fraction of total ASVs represented by that bar.

>>>>

---
title: "A Tale of Two Transects"
subtitle: "Also Two Sequencing Technologies, and Two Amplicons"
author: 
  - Brendan Furneaux^1^,
    Mohammad Bahram^1^,
    Anna Rosling^3^,
    Nourou S. Yorou^2^,
    Martin Ryberg^1^
  - ^1^Program in Systematic Biology,
    Department of Organismal Biology,
    Uppsala University,
    Uppsala, Sweden
  - ^2^Research Unit in Tropical Mycology and Plant-Fungi Interactions,
    LEB,
    University of Parakou,
    Parakou, Benin
  - ^3^Program in Evolutionary Biology,
    Department of Ecology and Genetics,
    Uppsala University,
    Uppsala, Sweden
date: "`r format(Sys.time(), '%B %d, %Y')`"
bibliography:
  - "all.bib"
  - "R.bib"
csl: molecular-ecology.csl
always_allow_html: true
output:
  bookdown::pdf_document2:
    fig_crop: yes
    keep_tex: yes
    latex_engine: lualatex
    includes:
      in_header: preamble.tex
  bookdown::word_document2: 
    keep_md: yes
---

```{r setup, include=FALSE}
library("knitr")
library("tikzDevice")
library("ggplot2")
library("magrittr")
library("tidyverse")
library("drake")
library("here")
library("phyloseq")
tryCatch(
  tinytex::tinytex_root(),
  error = function(e) tinytex::install_tinytex()
  )

knitr::opts_chunk$set(
  dev = "pdf",
  # dev.args = list(pointsize=32),
  echo = FALSE,
  # results = 'asis',
  # fig.show = 'asis',
  fig.width = 6,
  fig.height = 4.5,
  message = FALSE,
  warning = FALSE,
  external = TRUE
)
options(
  tikzDefaultEngine = "luatex",
  tikzXelatexPackages = c(options("tikzXelatexPackages")$tikzXelatexPackages,
                          "\\usepackage[version=4]{mhchem}\n",
                          "\\usepackage{siunitx}\n",
                          "\\DeclareSIUnit\\year{a}\n",
                          "\\usepackage{fontspec}"),
  tikzMetricsDictionary = here("writing/transect_paper-tikzDictionary")
)

@@

loadd(
  reads_table,
  region_table,
  asv_table,
  otu_table,
  demuxlength,
  demuxqual,
  readcounts,
  rawcounts,
  demuxcounts,
  regioncounts,
  cache = cache
)

@@

@@

Taxonomic assignment was performed to genus level separately on the ITS and LSU regions using Unite/Warcup and RDP, respectively, as taxonomic references.
For each region/reference combination, taxonomy was assigned using three algorithms:
the RDP Naïve Bayesian Classifier (RDPC) as implemented in DADA2;
SINTAX [@edgar2016a] as implemented in VSEARCH v2.9.1 [@rognes2016];
and IDTAXA [@murali2018].
Each ASV was thus given up to nine preliminary taxonomic assignments (three region/reference combinations $\times$ three algorithms).

@@

```{r bioinfo}
loadd(bioinf_table, cache = cache)
options(knitr.kable.NA = '–')
bioinf_header <- tibble(name = names(bioinf_table)) %>%
  separate(name, into = c("tech", "amplicon", "type"), sep = "_") %>%
  mutate(tech = paste(tech, plyr::mapvalues(tech, datasets$tech, datasets$machine, FALSE)))
bioinf_table %>%
  set_names(gsub(".*_", "", names(.))) %>%
  kable(
    booktabs = TRUE,
    caption = "(ref:readcounts)",
    align = "rrrrrr",
    format.args = list(big.mark = " "),
    linesep = c(rep("", 3), rep("\\addlinespace", 2), rep("", 4)),
    escape = FALSE
    ) %>%
  kableExtra::add_header_above(
    header = c(
      " " = 1,
      bioinf_header$amplicon[c(TRUE, FALSE)] %>% set_names(rep(2, length(.)), .)
    )
  ) %>%
  kableExtra::add_header_above(
    header = c(
      " " = 1,
      rle(bioinf_header$tech) %$% set_names(lengths, values)
    )
  )
```

@@

```{r venn-table-asv}
loadd(venn_ASV, cache = cache)

@@

```{r venn-table-otu}
loadd(venn_OTU, cache = cache)

@@

```{r pb-vs-it, fig.cap="(ref:readcap)", fig.height=8}
loadd(
  big_table,
  multi_table,
  comparisons,
  cache = cache
)

@@

@@

read_comparison <- function(big_table, comparisons, x, y, xlab, ylab) {
  x_var <- enquo(x)
  y_var <- enquo(y)
  big_table %>%
  filter(!!x_var > 0, !!y_var > 0) %>%
  ggplot(aes(x = !!x_var, y = !!y_var)) +
  geom_point(alpha = 0.2, shape = 1) +
  scale_x_log10() +
  scale_y_log10(limits = c(1, 1e4)) +
  coord_equal() +
  xlab(xlab) +
  ylab(ylab) +
  geom_rug(aes(y = !!y_var, x = 1), sides = "l", alpha = 0.2,
           data = big_table %>% filter(!!x_var == 0, !!y_var > 0)) +
  geom_rug(aes(y = !!x_var), sides = "b", alpha = 0.2,
           data = big_table %>% filter(!!y_var == 0, !!x_var > 0)) +
  stat_smooth(method = "loess", formula = y ~ x) +
  geom_abline(aes(intercept = mean(log10((!!y_var)[(!!y_var) >= 1])) -
                    mean(log10((!!x_var)[(!!x_var) >= 1])), slope = 1),
              linetype = 2L, alpha = 0.5) +
  geom_text(
    aes(label = label),
    data = filter(comparisons, .data$x == quo_name(x_var), .data$y == quo_name(y_var)),
    x = 1,
    y = 3.5,
    color = "black",
    hjust = 0,
    parse = TRUE,
    inherit.aes = FALSE
  ) +
  facet_wrap(~type, strip.position = "right", scales = "free")
}
```

@@

The fraction of reads and ASVs assigned to each taxonomic rank is shown in Figure \@ref(fig:tax-chart).
For all sequencing datasets and taxonomic assignment protocols, a higher proportion of reads was assigned than of ASVs, indicating that common ASVs were more likely to be identified than rare ASVs.
A greater fraction of ITS reads and ASVs were assigned using the Unite database than the Warcup database across sequencing technologies, amplicons, algorithms, and taxonomic ranks.
Across sequencing technologies, amplicons, and reference databases, the RDPC algorithm assigned the greatest fraction of reads and ASVs, followed by SINTAX, and then IDTAXA, with the exception of lower taxonomic ranks (order for the Long amplicon, family and genus for both amplicons) using the warcup reference database, where SINTAX assigned the greatest fraction of reads, but not ASVs.

(ref:tax-chart) Fraction of ASVs (left) and reads (right) assigned to each taxonomic rank, for different sequencing technologies (PacBio RS II, Ion Torrent Ion S5), amplicons (Long, Short), reference databases (Unite, Warcup, RDP), and assignment algorithms (PHYLOTAX, RDPC, SINTAX, IDTAXA).

```{r tax-chart, fig.cap = "(ref:tax-chart)"}
readd(taxon_reads, cache = cache) %>%
  pivot_longer(names_to = "rank", cols = c("kingdom", "phylum", "class", "order", "family", "genus"), values_to = "taxon") %>%
  group_by(amplicon, tech, Algorithm, reference, rank, ID = !is.na(taxon)) %>%
  summarize(reads = sum(reads), ASVs = n()) %>%
  group_by(amplicon, tech, Algorithm, reference, rank) %>%
  mutate(ASVs = ASVs / sum(ASVs)) %>%
  ungroup() %>%
  filter(ID) %>%
  mutate_at("rank", factor, levels = c("kingdom", "phylum", "class", "order", "family", "genus")) %>%
  rename(Reads = reads) %>%
  mutate_at("reference", as.character) %>%
  mutate(
    reference = map(reference, ~ if (. == "All") c("Unite", "Warcup", "RDP") else .)
  ) %>%
  unnest(reference) %>%
  mutate_at("reference", factor, levels = c("Unite", "Warcup", "RDP")) %>%
  pivot_longer(cols = c("Reads", "ASVs"), names_to = "type", values_to = "frac") %>%
  ggplot(aes(x = rank, y = frac, ymax = frac, group = Algorithm, fill = Algorithm)) +
  ggnomics::facet_nested(reference ~ type + tech + amplicon) +
  # geom_ribbon(ymin = 0, alpha = 0.5) +
  # geom_line(aes(color = Algorithm)) +
  geom_col(position = "dodge") +
  ylab("Fraction of reads assigned") +
  xlab(NULL) +
  theme(axis.text.x = element_text(vjust = 0.5, angle = 90),
        legend.direction = "horizontal",
        legend.position = "bottom")
```

Taxonomic composition of the sequenced soil community is summarized in Figure \@ref(fig:taxa).

@@

(ref:taxa-kingdom) Taxonomic assignment of ASVs to kingdom by different reference databases and algorithms.
"unidentified" denotes ASVs which could not be identified to kingdom level.
"other" denotes kingdoms represented by less than 1% of reads in all datasets.
The Warcup database did not include non-Fungi, so for that database, ASVs were considered to be identified to Fungi if they were identified to any phylum.
Height of each bar represents the fraction of reads represented by that bar.

```{r taxon-plot}
taxon_plot <- function(.data, rank, ..., y = reads, cutoff = NULL) {
  rank <- enquo(rank)
  y <- enquo(y)
  ranks <- c("kingdom", "phylum", "class", "order", "family", "genus")
  .data <- .data %>%
    group_by(reference, Algorithm, tech, amplicon) %>%
    mutate(ASVs = n()) %>%
    ungroup() %>%
    filter(...) %>%
    arrange_at(ranks) %>%
    mutate_at(ranks, ~ factor(., levels = c(NA, "other", discard(unique(.), is.na)))) %>%
    group_by(reference, Algorithm, tech, amplicon, !!rank) %>%
    summarize(reads = sum(reads), ASVs = n()/max(ASVs))

  if (!is.null(cutoff)) {
    .data <- group_by(.data, !!rank) %>%
      group_map(

        ~ if (all(pull(.x, !!y) < cutoff)) {
          mutate(.x, !!rank := factor("other", levels = levels(!!rank)))
        } else {
          .x
        },
        keep = TRUE
      ) %>%
      bind_rows() %>%
      group_by(reference, Algorithm, !!rank, tech, amplicon) %>%
      summarize(reads = sum(reads), ASVs = sum(ASVs)) %>%
      ungroup()
  }

  .data <- mutate(.data, !!rank := fct_drop(!!rank))
  vals <- levels(pull(.data, !!rank))
  # if ("other" %in% vals) vals <- c("other", vals) %>% magrittr::extract(!duplicated(.))
  # if (any(is.na(vals))) vals <- c(NA, vals) %>% magrittr::extract(!duplicated(.))
